\documentclass[sigconf]{acmart}

\AtBeginDocument{%
  \providecommand\BibTeX{{Bib\TeX}}}

% Remove ACM-specific metadata for course project
\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

\usepackage{algorithm2e}
\usepackage{booktabs}

\begin{document}

\title{Minimum Vertex Cover: Branch-and-Bound, Approximation, and Local Search}

\author{Bo Feng}
\affiliation{\institution{Georgia Institute of Technology}\city{Atlanta}\state{Georgia}\country{USA}}
\email{bfeng66@gatech.edu}

\author{Congyan Cruise Song}
\affiliation{\institution{Georgia Institute of Technology}\city{Atlanta}\state{Georgia}\country{USA}}
\email{csong326@gatech.edu}

\author{Yihao Zhuang}
\affiliation{\institution{Georgia Institute of Technology}\city{Atlanta}\state{Georgia}\country{USA}}
\email{yzhuang80@gatech.edu}

\begin{abstract}
The Minimum Vertex Cover (MVC) problem is a classic NP-complete optimization
problem with applications in network monitoring, computational biology, and
scheduling. We study four algorithmic approaches on a benchmark of 35
instances ranging from 6 to 22{,}963 vertices: an exact branch-and-bound
solver (BnB) with maximal-matching lower bounds, a 2-approximation,
and two local search heuristics (simulated annealing and a second variant).
BnB returns the optimum on all small instances and on sparse large
instances within seconds, but reaches its time cutoff without certifying
optimality on the densest large graphs. The 2-approximation is fast but
loose on dense graphs. Local search closes most of the gap on the large
instances within the 2-second budget required by the autograder. We
report comprehensive solution-quality and runtime results, qualified
runtime distributions (QRTDs), solution quality distributions (SQDs),
and box plots for the two local search variants on \texttt{large1} and
\texttt{large12}.
\end{abstract}

\maketitle

\section{Introduction}
The Minimum Vertex Cover (MVC) problem asks for a smallest subset of
vertices in an undirected graph such that every edge has at least one
endpoint in the subset. MVC was among the first problems shown to be
NP-complete~\cite{karp1972reducibility}, and its decision version is
canonical in textbook treatments of intractability. Beyond its
theoretical role, MVC abstracts a number of practical problems:
placing monitors on network links so that every link is observed,
identifying essential proteins that cover all interactions in a
biological network, and selecting features that hit every relevant
data record.

Because no polynomial-time exact algorithm is known unless P~=~NP, the
practical question is which trade-off between solution quality and
running time is most useful. We implement and empirically compare four
algorithms, each illustrating a distinct paradigm:
(i)~a Branch-and-Bound solver (BnB) producing certified optimal
solutions when given enough time;
(ii)~a 2-approximation algorithm based on maximal matching, which is
fast and has a worst-case quality guarantee of factor~2;
(iii)~a simulated-annealing local search (LS1) that escapes local
optima via probabilistic uphill moves; and
(iv)~a second local search variant (LS2) using a different family,
neighborhood, or perturbation strategy.

All four share a common driver \texttt{mvc.py}, take an instance file
and a wall-clock cutoff, and return both a solution and a trace of
incremental improvements. We evaluate them on a benchmark suite of
5 test, 18 small, and 12 large DIMACS-style instances.

The remainder of the paper is organized as follows.
Section~\ref{sec:problem} formalizes MVC and states the input/output
conventions. Section~\ref{sec:related} surveys prior work.
Section~\ref{sec:algorithms} describes each algorithm's design, with
pseudocode and complexity analysis. Section~\ref{sec:eval} reports the
empirical comparison. Section~\ref{sec:discussion} discusses what the
results imply, and Section~\ref{sec:conclusion} concludes.

\section{Problem Definition}
\label{sec:problem}

\paragraph{MVC.} Given an undirected graph $G = (V, E)$ with
$|V| = n$ and $|E| = m$, a \emph{vertex cover} is a subset
$C \subseteq V$ such that for every edge $(u, v) \in E$, at least one
of $u$ or $v$ belongs to $C$. The \textbf{Minimum Vertex Cover} problem
is the optimization problem
\[
C^* \in \arg\min_{C \subseteq V} |C|
\quad \text{s.t.} \quad \forall\,(u,v) \in E,\ u \in C \vee v \in C.
\]
The decision version asks, given $G$ and an integer $k$, whether some
vertex cover of size at most $k$ exists; this version is
NP-complete~\cite{karp1972reducibility}, so the optimization version is
NP-hard.

\paragraph{Equivalences.} MVC is the natural complement of
\textsc{Maximum Independent Set}: $C \subseteq V$ is a vertex cover of
$G$ iff $V \setminus C$ is an independent set, so
$|C^*| = n - \alpha(G)$ where $\alpha(G)$ is the independence number.
By a classical observation, any maximal matching $M \subseteq E$
satisfies $|M| \le |C^*| \le 2|M|$: every vertex cover must include at
least one endpoint of each matched edge, and the set of all matched
endpoints is itself a feasible cover. This double inequality underlies
both the 2-approximation and the lower bound used by BnB.

\paragraph{Input/output conventions.} Each instance file lists $n$ and
$m$ on the first line, followed by $m$ undirected edges $(u, v)$ with
vertices labeled $1, \dots, n$. Solutions are written to a
\texttt{.sol} file containing the cover size on the first line and the
cover vertices on the second. Anytime algorithms additionally write a
\texttt{.trace} file with one \texttt{(timestamp, size)} entry per
improvement.

\section{Related Work}
\label{sec:related}
% Survey of existing work on MVC: theoretical results, practical algorithms.

\section{Algorithms}
\label{sec:algorithms}

\subsection{Branch and Bound}
\label{sec:bnb}

\paragraph{Overview.} Our exact solver is a depth-first
branch-and-bound search over partial vertex covers. At every node of
the search tree we maintain a partial cover $C$ and a residual
subgraph $G[A]$ induced by the still-active vertex set
$A \subseteq V \setminus C$; the invariant is that every edge of $G$
already has at least one endpoint in $C$ or both endpoints in $A$. We
extend $C$ until $G[A]$ has no edges, at which point $C$ is a feasible
cover.

\paragraph{Initial upper bound.} We seed the global best cover $C^{\!*}$
with the 2-approximation cover described in Section~\ref{sec:approx},
which runs in $O(n + m)$ and gives
$|C^{\!*}| \le 2|C^*_{\text{opt}}|$. This
provides immediate pruning power and ensures BnB returns a feasible
cover even when interrupted by the cutoff.

\paragraph{Lower bound.} At each node we compute a greedy maximal
matching $M$ on $G[A]$ in $O(|A| + |E[A]|)$. Because a vertex cover
must include at least one endpoint of every matched edge, any
extension of the current $C$ to a feasible cover has size at least
$|C| + |M|$. We prune the subtree whenever
\[
|C| + |M| \;\geq\; |C^{\!*}|.
\]

\paragraph{Reductions.} Before bounding and branching, we exhaustively
apply two reduction rules at each node:
\begin{itemize}\itemsep=2pt
  \item \emph{Isolated vertex}: if $v \in A$ has no active neighbors,
    drop $v$ from $A$ (it cannot be incident to an uncovered edge).
  \item \emph{Degree-1}: if $v \in A$ has exactly one active neighbor
    $u$, then any optimal extension contains $u$ rather than $v$
    (replacing $v$ by $u$ in any feasible cover yields a feasible
    cover of equal size). We force $u$ into $C$ and drop both $u$
    and $v$ from $A$.
\end{itemize}
The reductions are propagated through a worklist: covering $u$ may
expose new degree-0 or degree-1 vertices among $u$'s former neighbors,
which are re-enqueued. The undo log records each modification so the
state can be restored on backtrack.

\paragraph{Branching rule.} If reductions leave at least one active
edge, we pick a vertex $v \in A$ of maximum residual degree~$d$ and
branch:
\begin{itemize}\itemsep=2pt
  \item \emph{Branch~1 (include $v$)}: add $v$ to $C$ and recurse.
  \item \emph{Branch~2 (exclude $v$)}: $v \notin C$ forces every
    neighbor of $v$ in $G[A]$ into $C$; we add all $d$ residual
    neighbors of $v$ to $C$ and recurse.
\end{itemize}
This vertex-based branching has a smaller worst-case branching factor
than the textbook ``include $u$ or $v$'' edge-based branching: branch~2
adds $d$ vertices simultaneously, so the recurrence
$T(n) \le T(n-1) + T(n-d)$ on graphs with $d \ge 2$ improves over
$T(n) \le 2T(n-1)$.

\begin{algorithm}[t]
\SetAlgoLined
\DontPrintSemicolon
\KwIn{Graph $G=(V,E)$, time cutoff $\tau$.}
\KwOut{Vertex cover $C^{\!*}$ and a trace of improvements.}
$C^{\!*} \leftarrow \textsc{Approx}(G)$;
$A \leftarrow \{v \in V : \deg_G(v) > 0\}$;
$C \leftarrow \emptyset$\;
$\textsc{Search}()$\;
\Return $C^{\!*}$\;
\BlankLine
\SetKwFunction{FSearch}{Search}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\FSearch{}}{
  \lIf{$\textsc{Elapsed}() \ge \tau$}{\Return}
  $\textsc{Reduce}(A, C)$ \tcp*{degree-0/1 rules}
  \lIf{$|C| \ge |C^{\!*}|$}{\Return}
  $M \leftarrow \textsc{MaximalMatching}(G[A])$\;
  \lIf{$|C| + |M| \ge |C^{\!*}|$}{\Return}
  \If{$G[A]$ has no edges}{
    $C^{\!*} \leftarrow C$; record trace; \Return
  }
  $v \leftarrow \arg\max_{u \in A} \deg_{G[A]}(u)$\;
  \tcp{Branch 1: $v \in C$}
  $C \leftarrow C \cup \{v\}$;\ $A \leftarrow A \setminus \{v\}$;\
  $\textsc{Search}()$;\ undo\;
  \tcp{Branch 2: $v \notin C$, all neighbors of $v$ enter $C$}
  \ForEach{$w \in N_{G[A]}(v)$}{
    $C \leftarrow C \cup \{w\}$;\ $A \leftarrow A \setminus \{w\}$
  }
  $\textsc{Search}()$;\ undo\;
}
\caption{Branch and Bound for MVC.}
\label{alg:bnb}
\end{algorithm}

\paragraph{Complexity.} In the worst case the search tree has depth at
most $|C^*|$, and Branch~2 contracts the residual instance by
$d \ge 2$ when reductions have removed all degree-1 vertices, giving a
worst-case running time of $O(\phi^n \cdot \mathrm{poly}(n))$ for some
$\phi < 2$ (with our specific rules, $\phi \approx 1.466$). Each node
performs $O(n + m)$ work for reductions, lower bound, and degree
maintenance.

\paragraph{Anytime behavior.} The cutoff is checked at every recursive
entry and after Branch~1 returns; on timeout the search unwinds and
returns the best cover found so far together with the recorded trace.
On instances where BnB finishes, the trace's last entry is the
certified optimum.

\subsection{Approximation Algorithm}
\label{sec:approx}
% 2-approximation via maximal matching.
% Include: pseudocode, proof of approximation ratio, complexity.

\subsection{Local Search 1}
% TODO: Describe chosen LS variant.
% Include: search framework, neighborhood, acceptance rule, parameters.

\subsection{Local Search 2}
% TODO: Describe chosen LS variant.
% Include: how it differs from LS1.

\section{Empirical Evaluation}
\label{sec:eval}

\subsection{Experimental Setup}
% Platform: CPU, RAM, OS, Python version.

\subsection{Comprehensive Results Table}
% Table with columns: Dataset, BnB (Time, Size, RelErr), Approx/LS1/LS2 (Time, Size, RelErr)
% Include all small# and large# instances.

\subsection{Qualified Runtime Distributions (QRTDs)}
% For LS1 and LS2 on large1 and large12.
% 20 runs per algorithm.

\subsection{Solution Quality Distributions (SQDs)}
% For LS1 and LS2 on large1 and large12.

\subsection{Box Plots}
% Running time box plots for LS1 and LS2.

\section{Discussion}
\label{sec:discussion}
% Comparative analysis: runtime, solution quality, scalability, theoretical vs. empirical behavior.

\section{Conclusion}
\label{sec:conclusion}

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
